Patient-Specific Neural Field Modeling from Micro-Electrode Data

### 2010-08-18 ltx/data_paper/DataPaper.tex
corrected the latex file
Fixed error in previous upload
added observation noise estimation

To show that this is indeed the case consider the function $\kappa(v(\mathbf{r},t),t) = v(\mathbf{r},t)e^{\zeta t}$. We note that $\kappa(v(\mathbf{r},t),t)$ is twice differentiable so that we can apply It\hat{o}'s formula to obtain

@@

\subsection{$\mathcal Q$-function}
The goal of the M-step is to maximise the likelihood function $p(\mathbf Y;\boldsymbol \theta,\xi)$ given the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)$ over the observed variables $\mathbf Y$ and hidden variables $\mathbf X$ that is governed by parameters $\boldsymbol \theta$ and $\xi$. Unfortunately, it is often not possible to maximise the likelihood $p(\mathbf Y;\boldsymbol \theta,\xi)$ directly. However, by using the relationship between $p(\mathbf Y;\boldsymbol \theta,\xi)$ and $p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)$ it is possible to compute a sequence of parameter estimates that converges on the maximum likelihood parameter estimate.

Expanding the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)$ we have

@@

 p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)=\prod_{t=0}^{T-1} p(\mathbf y_{t+1}|\mathbf x_{t+1})p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta, \xi)p(\mathbf x_0).

@@

 \mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')&=\mathbf E_{\boldsymbol (\theta',\xi')}\left[\ln p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)\right] \\
&=\mathbf E_{\boldsymbol (\theta',\xi')}\left[\sum_{t=0}^{T-1}\ln p(\mathbf y_t|\mathbf x_t)+\sum_{t=0}^{T-1}\ln p(\mathbf x_t|\mathbf x_{t-1};\boldsymbol \theta, \xi)+\sum_{t=0}^{T-1}\ln p(\mathbf x_0)\right],

@@

where $ \mathbf E_{\boldsymbol (\theta',\xi')}$ denotes expectation taken with respect to the distribution $p(\mathbf X | \mathbf Y;\boldsymbol \theta',\xi') $ where $\boldsymbol \theta'$ and $\xi'$are the current parameter estimates. It should be noted that
$p(\mathbf y_t|\mathbf x_t)$ and $p(\mathbf x_0)$ are not functions of parameter set and therefore $\mathcal Q$-function can be rewritten as

@@

\mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')=\mathbf E_{(\theta',\xi')}\left[\sum_{t=0}^{T-1}\ln p(\mathbf x_t|\mathbf x_{t-1};\boldsymbol \theta ,\xi)\right]+\mathrm{const}

@@

the constant term can be ignored since it is not affecting the derivative with respect to the parameters. 
The Gaussian distribution of the disturbance, $e_t(\mathbf r)\sim \mathcal N(\mathbf 0,\boldsymbol\Sigma_e(\mathbf r,\mathbf r'))$ results into the following conditional distribution of the state at $t+1$
 p(\mathbf x_{t+1} | \mathbf x_t;\boldsymbol\theta,\xi)=\alpha e^{-(\mathbf x_{t+1}-q(\mathbf  x_t)\boldsymbol\theta-\xi  \mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}(\mathbf x_{t+1}-q( \mathbf x_t)\boldsymbol\theta-\xi \mathbf  x_t)}
where $\alpha$ is the normalising constant. By expanding the exponent and taking $\log$ we get

@@

  \ln p(\mathbf x_{t+1} | \mathbf x_t;\boldsymbol\theta,\xi)&=&\ln \alpha-\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}+2\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}q( \mathbf x_t)\boldsymbol\theta+2\xi \mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_t\nonumber \\
&&-\boldsymbol\theta^\top q^\top(\mathbf x_t)\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-2\xi \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-\xi^2\mathbf x_t^\top\boldsymbol\Sigma_e^{-1}\mathbf x_t.

@@

  \ln p(\mathbf x_{t+1} | \mathbf x_t;\boldsymbol\theta,\xi)&=&\ln \alpha-\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}+2\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}q( \mathbf x_t)\boldsymbol\theta+2\xi \mathrm{tr} \left\lbrace \mathbf x_t\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\right\rbrace \nonumber \\

@@

\begin{align}\label{eq:Voldermont}
 \mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')&\approx\beta+2\Xi_0\boldsymbol\theta+
2\xi tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace \nonumber \\
&-\boldsymbol\theta^\top \Xi_2\boldsymbol\theta-2\xi \Xi_3 \boldsymbol\theta - \xi^2 tr \left\lbrace\Xi_4\boldsymbol\Sigma_e^{-1} \right\rbrace \tag{Voldermont}

@@

	\beta &= \sum_{t=0}^{T-1}\mathbf E_{(\theta',\xi')}\left[ \ln \alpha-\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}\right] \label{eq:def of beta} \\

@@

\Xi_1&=\mathbf E_{(\theta',\xi')}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\

@@

 \Xi_4&=\mathbf E_{(\theta',\xi')}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right)

@@

Given the state estimates, $\mathbf {\hat x^{b}}_t$ the state covariance $\mathbf P_t^b$ and cross-covariance $\mathbf M_t^b$ , the $\Xi$-variables can be computed directly.

When taking the derivative of equation~\ref{eq:Voldermont}, the first term described in equation~\ref{eq:def of beta} will disappear. Differntiating with respect to $\boldsymbol\theta$ we have
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=2\Xi_0-\boldsymbol\theta^\top(\Xi_2^\top+\Xi_2)-2\xi\Xi_3
\end{equation}
noting that $\Xi_2$ is symmetric and equating \ref{eq:Qpartialtheta} to the zero vector we get
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\Xi_2^{-1}\left(\Xi_0^\top-\xi\Xi_3^\top \right)
\end{equation}
similarly differentiating with respect to $\xi$ we have
\begin{equation}\label{eq:Qpartialxi}
 \frac{\partial \mathcal Q}{\partial \xi}=2tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace -2 \Xi_3 \boldsymbol\theta - 2\xi tr \left\lbrace\Xi_4\boldsymbol\Sigma_e^{-1} \right\rbrace
\end{equation}
substituting \ref{eq:theta} in \ref{eq:Qpartialxi} and equating to zero we get
\begin{equation}
 \xi=\frac{tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_0^\top}{tr\left\lbrace \Xi_4 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_3^\top}
\end{equation}
\subsection{Discussion}
The second derivative of the $\mathcal Q$-function with respect to $\xi$ is given by
\begin{equation}
 \frac{\partial^2 \mathcal Q}{\partial \xi^2}=-2 tr \left\lbrace\Xi_1\boldsymbol\Sigma_e^{-1} \right\rbrace
\end{equation}
$\Xi_1 $ is positive definite if the state vector is persistently excited and the covariance matrix $\boldsymbol\Sigma_e^{-1}$ is also positive definite therefore the trace of $\Xi_1\boldsymbol\Sigma_w^{-1}$ is guaranteed to be positive so the second derivative is negative and represents a maximum of $\mathcal Q$-function.\\
The second derivative of the $\mathcal Q$-function with respect to $\boldsymbol \theta$ is given by
\begin{equation}
 \frac{\partial^2 \mathcal Q}{\partial \boldsymbol\theta^2}=-2 \Xi_2 
\end{equation}
$\Xi_2$ is positive definite and invertible and the second derivative is negative definite and represents a maximum of $\mathcal Q$- function.

>>>>

To show that this is indeed the case consider the function $\kappa(v(\mathbf{r},t),t) = v(\mathbf{r},t)e^{\zeta t}$. We note that $\kappa(v(\mathbf{r},t),t)$ is twice differentiable so that we can apply Ito's formula to obtain

@@

\subsection{$\mathcal Q$-function}
The goal of the M-step is to maximise the likelihood function $p(\mathbf Y;\boldsymbol \Theta)$ given the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$ over the observed variables $\mathbf Y$ and hidden variables $\mathbf X$ that is governed by parameters $\boldsymbol\Sigma_{\epsilon}$, $\boldsymbol \theta$ and $\xi$. Unfortunately, it is often not possible to maximise the likelihood $p(\mathbf Y;\boldsymbol\Theta)$ directly. However, by using the relationship between $p(\mathbf Y;\boldsymbol \Theta)$ and $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$ it is possible to compute a sequence of parameter estimates that converges on the maximum likelihood parameter estimate.

Expanding the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$ we have

@@

 p(\mathbf X,\mathbf Y;\boldsymbol \Theta)=\prod_{t=0}^{T-1} p(\mathbf y_{t+1}|\mathbf x_{t+1};\boldsymbol \Sigma_{\epsilon})p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta, \xi)p(\mathbf x_0).

@@

 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&=\mathbf E_{\boldsymbol \Theta'}\left[2\ln p(\mathbf X,\mathbf Y;\boldsymbol \Theta)\right] \\
&=\mathbf E_{\boldsymbol\Theta'}\left[\sum_{t=0}^{T-1}2\ln p(\mathbf y_{t+1}|\mathbf x_{t+1};\boldsymbol \Sigma_{\epsilon})+\sum_{t=0}^{T-1}2\ln p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta, \xi)+\sum_{t=0}^{T-1}2\ln p(\mathbf x_0)\right],

@@

where $ \mathbf E_{\boldsymbol \Theta'}$ denotes expectation taken with respect to the distribution $p(\mathbf X | \mathbf Y;\boldsymbol \Theta') $ where $\boldsymbol \Theta'$ is the current parameter estimate. It should be noted that
 $p(\mathbf x_0)$ is not function of parameter set and therefore $\mathcal Q$-function can be rewritten as

@@

\mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}2\ln p(\mathbf y_{t+1}|\mathbf x_{t+1};\boldsymbol \Sigma_{\epsilon})+\sum_{t=0}^{T-1}2\ln p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\xi)\right]+\mathrm{const}

@@

the constant term can be ignored since it is not affecting the derivative with respect to the parameters. Under the Gaussian assumption and ignoring constant normalising terms, the joint distribution at $t+1$ can be written as
\begin{eqnarray}
 p(\mathbf x_{t+1}, \mathbf y_{t+1};\boldsymbol\Theta)&=& \mid\boldsymbol\Sigma_{\epsilon}\mid^{-\frac{1}{2}} \times  e^{-\frac{1}{2}(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1})^\top\boldsymbol\Sigma_{\epsilon}^{-1}(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1}) } \nonumber \\
&+& e^{-\frac{1}{2}(\mathbf x_{t+1}-q(\mathbf  x_t)\boldsymbol\theta-\xi  \mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}(\mathbf x_{t+1}-q( \mathbf x_t)\boldsymbol\theta-\xi \mathbf  x_t)}
by expanding the exponent and taking $2\log$ we get

@@

2\ln p(\mathbf x_{t+1} , \mathbf y_{t+1};\boldsymbol\Theta)&=&-\ln \boldsymbol\Sigma_{\epsilon}- (\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1})^\top\boldsymbol\Sigma_{\epsilon}^{-1}(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1})\nonumber \\
&-&\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}+2\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}q( \mathbf x_t)\boldsymbol\theta+2\xi \mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_t\nonumber \\
&-&\boldsymbol\theta^\top q^\top(\mathbf x_t)\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-2\xi \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-\xi^2\mathbf x_t^\top\boldsymbol\Sigma_e^{-1}\mathbf x_t.

@@

  \ln p(\mathbf x_{t+1}, \mathbf y_{t+1};\boldsymbol\Theta)&=& -\ln \boldsymbol\Sigma_{\epsilon}-\mathrm{tr}\left\lbrace\boldsymbol\Sigma_{\epsilon}^{-1}(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1})^\top\right\rbrace   \nonumber \\
&-&\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}+2\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}q( \mathbf x_t)\boldsymbol\theta+2\xi \mathrm{tr} \left\lbrace \mathbf x_t\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\right\rbrace \nonumber \\

@@

\begin{eqnarray}\label{eq:Voldermont}
 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&\approx& -(T-1)\ln \boldsymbol\Sigma_{\epsilon}-\mathrm{tr}\left\lbrace\boldsymbol\Sigma_{\epsilon}^{-1}\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\
&+&\beta+2\Xi_0\boldsymbol\theta+
2\xi \mathrm{tr}\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace \nonumber \\
&-&\boldsymbol\theta^\top \Xi_2\boldsymbol\theta-2\xi \Xi_3 \boldsymbol\theta - \xi^2 \mathrm{tr} \left\lbrace\Xi_4\boldsymbol\Sigma_e^{-1} \right\rbrace 

@@

	\beta &= \sum_{t=0}^{T-1}\mathbf E_{\Theta'}\left[-\mathbf x_{t+1}^\top\boldsymbol\Sigma_e^{-1}\mathbf x_{t+1}\right] \label{eq:def of beta} \\

@@

\Xi_1&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t+1}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\

@@

 \Xi_4&=\mathbf E_{\Theta'}\left[\sum_{t=0}^{T-1}\mathbf x_t\mathbf x_{t}^\top\right]=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right)

@@

It should be noted that the second term in equation \ref{eq:Voldermont} is obtained by adding and subtracting $\mathbf C \mathbf{\hat{x}}_t \mathbf{\hat{x}}_t^\top \mathbf C^\top$ to the expectation, $\mathbf E_{\boldsymbol\Theta'}\left[ .\right] $. Given the state estimates, $\mathbf {\hat x^{b}}_t$ the state covariance $\mathbf P_t^b$ and cross-covariance $\mathbf M_t^b$ , the $\Xi$-variables can be computed directly.

When taking the derivative of equation~\ref{eq:Voldermont}, the constant $\beta$ described in equation~\ref{eq:def of beta} will disappear. Differntiating with respect to $\boldsymbol\theta$ we have
\begin{equation}\label{eq:Qpartialtheta}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\theta}=2\Xi_0-\boldsymbol\theta^\top(\Xi_2^\top+\Xi_2)-2\xi\Xi_3
\end{equation}
noting that $\Xi_2$ is symmetric and equating \ref{eq:Qpartialtheta} to the zero vector we get
\begin{equation}\label{eq:theta}
 \boldsymbol \theta=\Xi_2^{-1}\left(\Xi_0^\top-\xi\Xi_3^\top \right)
\end{equation}
similarly differentiating with respect to $\xi$ we have
\begin{equation}\label{eq:Qpartialxi}
 \frac{\partial \mathcal Q}{\partial \xi}=2tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace -2 \Xi_3 \boldsymbol\theta - 2\xi tr \left\lbrace\Xi_4\boldsymbol\Sigma_e^{-1} \right\rbrace
\end{equation}
substituting \ref{eq:theta} in \ref{eq:Qpartialxi} and equating to zero we get
\begin{equation}
 \xi=\frac{tr\left\lbrace \Xi_1 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_0^\top}{tr\left\lbrace \Xi_4 \boldsymbol\Sigma_e^{-1}\right\rbrace-\Xi_3\Xi_2^{-1}\Xi_3^\top}
\end{equation}
differentiating equation~\ref{eq:Voldermont} with respect to $\boldsymbol\Sigma_{\epsilon}$ we get
\begin{equation}
 \frac{\partial \mathcal Q}{\partial \boldsymbol\Sigma_{\epsilon}}= (T-1)\boldsymbol\Sigma_{\epsilon}-\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right]=\mathbf 0
\end{equation}
therefore we have
\begin{equation}
 \boldsymbol\Sigma_{\epsilon}=\frac{1}{T-1}\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right]
\end{equation}

\subsection{Discussion}
The second derivative of the $\mathcal Q$-function with respect to $\xi$ is given by
\begin{equation}
 \frac{\partial^2 \mathcal Q}{\partial \xi^2}=-2 tr \left\lbrace\Xi_1\boldsymbol\Sigma_e^{-1} \right\rbrace
\end{equation}
$\Xi_1 $ is positive definite if the state vector is persistently excited and the covariance matrix $\boldsymbol\Sigma_e^{-1}$ is also positive definite therefore the trace of $\Xi_1\boldsymbol\Sigma_w^{-1}$ is guaranteed to be positive so the second derivative is negative and represents a maximum of $\mathcal Q$-function.\\
The second derivative of the $\mathcal Q$-function with respect to $\boldsymbol \theta$ is given by
\begin{equation}
 \frac{\partial^2 \mathcal Q}{\partial \boldsymbol\theta^2}=-2 \Xi_2 
\end{equation}
$\Xi_2$ is positive definite and invertible and the second derivative is negative definite and represents a maximum of $\mathcal Q$- function.
